Equivalence Testing and the Second Generation \emph{P}-Value

### 2018-08-28 manuscript.Rmd
submission

abstract: |
  To move beyond the limitations of null-hypothesis tests, statistical approaches have been developed where the observed data is compared against a range of values that are equivalent to the absence of a meaningful effect. Specifying a range of values around zero allows researchers to statistically reject the presence of effects large enough to matter, and prevents practically insignificant effects from being interepreted as a statistically significant difference. We compare the behavior of the recently proposed second generation *p*-value [@blume_second-generation_2018] with the more established Two One-Sided Tests (TOST) equivalence testing procedure [@schuirmann_comparison_1987]. We show that the two approaches yield almost identical results under optimal conditions. Under suboptimal conditions (e.g., when the confidence interval is wider than the equivalence range, or when confidence intervals are asymmetric) the second generation *p*-value becomes difficult to interpret as a descriptive statistic. The second generation *p*-value is interpretable in a dichotomous manner (i.e., when the SGPV equals 0 or 1 because the confidence intervals lies completely within or outside of the equivalence range), but this dichotomous interpretation does not require calculations. We conclude that equivalence tests yield more consistent *p*-values, distinguish between datasets that yield the same second generation *p*-value, and allow for easier control of Type I and Type II error rates.

@@

The TOST procedure also relies on the confidence interval around the effect. 
In the TOST procedure the data is tested against the lower equivalence bound in the first one-sided test, and against the upper equivalence bound in the second one-sided test [@lakens_equivalence_2018]. 
If both tests statistically reject an effect as extreme or more extreme than the equivalence bound, you can conclude the observed effect is practically equivalent to zero from a Neyman-Pearson approach to statistical inferences. 
Because one-sided tests are performed, one can also conclude equivalence by checking whether the 1-2$\times$alpha confidence interval (e.g., when the alpha level is 0.05, a 90% CI) falls completely within the equivalence bounds. 
Because both equivalence tests as the SGPV are based on whether and how much a confidence interval overlaps with equivalence bounds, it seems worthwhile to compare the behavior of the newly proposed SGPV to equivalence tests to examine the unique contribution of the SGPV to the statistical toolbox.

@@

The second generation *p*-value (SGPV) is calculated as:
$$
  p _ { \delta } = \frac { \left| I \cap H _ { 0 } \right| } { | I | } \times \max \left\{ \frac { | I | } { 2 \left| H _ { 0 } \right| } , 1 \right\}
$$
where I is the interval based on the data (e.g., a 95% confidence interval) and H~0~ is the equivalence range. The first term of this formula implies that the second generation *p*-value is the width of the confidence interval that overlaps with the equivalence range, divided by the total width of the confidence interval. The second term is a 'small sample correction' (which will be discussed later) that comes into play whenever the confidence interval is more than twice as wide as the equivalence range.
To examine the relation between the TOST *p*-value and the SGPV we can calculate both statistics across a range of observed effect sizes. In Figure \ref{fig:TOSTSGPV1} *p*-values are plotted for the TOST procedure and the SGPV. The statistics are calculated for hypothetical one-sample *t*-tests for observed means ranging from 140 to 150 (on the x-axis). The equivalence range is set to 145 ± 2 (i.e., an equivalence range from 143 to 147), the observed standard deviation is assumed to be 2, and the sample size is 30. For example, for the left-most point in Figure \ref{fig:TOSTSGPV1} the SGPV and the TOST *p*-value is calculated for a hypothetical study with a sample size of 30, an observed standard deviation of 2, and an observed mean of 140, where the *p*-value for the equivalence test is 1, and the SGPV is 0. Our conclusions about the relationship between TOST *p*-values and SGPV in this article are not dependent upon any specific example, as readers can explore for themselves in an online Shiny app: http://shiny.ieis.tue.nl/TOST_vs_SGPV/.
```{r, include=FALSE}
step = 0.01

@@

tost_res2 <- TOSTone.raw(m = m, 
                   mu = mu,
                   sd = sd, 
                   n = n, 
                   low_eqbound = low_eqbound, 
                   high_eqbound = high_eqbound, 
                   alpha = alpha,
                   plot = FALSE,
                   verbose = FALSE
)
SGPV2 <- TOST_to_SGPV(tost_res = tost_res2)
```
When the observed sample mean is 145, the sample size is 30, and the standard deviation is 2, and we are testing against equivalence bounds of 143 and 147 using the TOST procedure for a one-sample *t*-test, the equivalence test is significant, *t*(`r tost_res1$TOST_df`) = `r round(tost_res1$TOST_t1, digits = 2)`, *p* < .001. Because the 95% CI falls completely within the equivalence bounds, the SGPV is `r SGPV1` (see Figure \ref{fig:TOSTSGPV1}). 
On the other hand, when the observed mean is 140, the equivalence test is not significant (the observed mean is far outside the equivalence range of 143 to 147), *t*(`r tost_res2$TOST_df`) = `r round(tost_res2$TOST_t1, digits = 2)`, *p* = `r round(tost_res2$TOST_p1, digits =  4)` (or more accurately, *p* > .999 as *p*-values are bounded between 0 and 1). Because the 95% CI falls completely outside the equivalence bounds, the SGPV is `r SGPV2` (see Figure  \ref{fig:TOSTSGPV1}).

@@

Two other points always have to overlap. When the 95% CI falls completely inside the equivalence region, and one endpoint of the confidence interval is exactly equal to one of the equivalence bounds (see situation B in Figure \ref{fig:TOSTSGPV3}) the TOST *p*-value (which relies on a one-sided test) is always 0.025. The third point where the SGPV and the *p*-value from the TOST procedure should overlap is where the 95% CI falls completely outside of the equivalence range, but one endpoint of the confidence interval is equal to the equivalence bound (see situation C in Figure \ref{fig:TOSTSGPV3}), when the *p*-value will always be 0.975. Note that this situation is in essence a *minimum-effect test* [@murphy_statistical_2014]. The goal of a minimum-effect is not just to reject a difference of zero, but to reject the smallest effect size of interest (i.e., the equivalence bounds). The SGPV summarizes the information from an equivalence test and a minimum-effect test. These can be two relevant questions to ask, although it often makes sense to combine an equivalence test and a null-hypothesis test instead [@lakens_equivalence_2018].

@@

tost_res4 <- TOSTone.raw(m = m, 
                   mu = mu,
                   sd = sd, 
                   n = n, 
                   low_eqbound = low_eqbound, 
                   high_eqbound = high_eqbound, 
                   alpha = alpha,
                   plot = FALSE,
                   verbose = FALSE
)
SGPV4 <- TOST_to_SGPV(tost_res = tost_res4)
```
```{r , echo=FALSE}
tail1 <- 1-pnorm(2, 1.5, 0.5)
tail2 <- 1-pnorm(2, 1.4, 0.5)
tail3 <- 1-pnorm(2, 1.3, 0.5)
tail4 <- 1-pnorm(2, 1.2, 0.5)
```
For example, the SGPV from A to D is `r round(SGPV1, digits = 2)`, `r round(SGPV2, digits = 2)`, `r round(SGPV3, digits = 2)`, and `r round(SGPV4, digits = 2)`. The difference in the percentage of overlap between A and B (`r round(SGPV1-SGPV2, digits = 2)`) is identical to the difference in the percentage of overlap between C and D as the mean gets 0.1 closer to the test value (`r round(SGPV3-SGPV4, digits = 2)`).
As the observed mean in a one-sample *t*-test lies closer to the test value (in Figure \ref{fig:TOSTSGPV4}), from situation A to D, the mean gets closer to the test value by 0.1) the difference in the overlap changes uniformly. 
As we move the observed mean closer to the test value in steps of 0.1 across A to D the *p*-value calculated for normally distributed data is not uniformly distributed. The probability of observing data more extreme than the upper bound of 2 is (from A to D) `r round(tail1, digits = 3)`, `r round(tail2, digits = 3)`, `r round(tail3, digits = 3)`, and `r round(tail4, digits = 3)`. As we can see, the difference between A and B (`r round(tail1-tail2, digits = 3)`) is not the same as the difference between C and D (`r round(tail3-tail4, digits = 3)`). Indeed, the difference in *p*-values is the largest as you start at *p* = 0.5 (when the observed mean falls on the test value), which is why the line in Figure \ref{fig:TOSTSGPV1} is the steepest at *p* = 0.5. Note that where the SGPV reaches 1 or 0, *p*-values closely approximate 0 and 1, but never reach these values.

@@

There are three situations where *p*-values from TOST and SGPV are unrelated. The first two situations were discussed before and can be seen in Figure \ref{fig:TOSTSGPV1}. When the SGPV is either 0 or 1 *p*-values from the equivalence test fall between 0.975 and 1 or between 0 and 0.025. Because *p*-values approach 0 or 1, but are never exactly 0 or 1, and the SGPV is exactly 0 or 1, the two statistics are completely unrelated within these ranges. The easiest way to see this is by plotting the SGPV against the *p*-value from the TOST procedure. The situations where the SPGV and *p*-values from the TOST procedure are unrelated are indicated by the parts of the curve where there are vertical straight lines at second generation *p*-values of 0 and 1.
```{r, include=FALSE}
step = 0.01

@@

  res <- TOSTone.raw(m = m, 
                     mu = mu,
                     sd = sd, 
                     n = n, 
                     low_eqbound = low_eqbound, 
                     high_eqbound = high_eqbound, 
                     alpha = alpha,
                     plot = FALSE,
                     verbose = FALSE
  )
  t <- (m - mu)/(sd/sqrt(n))
  t_list[count] <- t
  sgpv_list[count] <- p_delta(mu+res$LL_CI_TTEST, mu+res$UL_CI_TTEST, mu+low_eqbound, mu+high_eqbound)
  p_tost_list[count] <- max(res$TOST_p1, res$TOST_p2)
  p_list[count] <- 2 * pt(-abs(t), df = n-1)
}
```
```{r, TOSTSGPV5, fig.cap="The relationship between *p*-values from the TOST procedure and the SGPV for the same scenario as in Figure \\ref{fig:TOSTSGPV1}."}
plot(sgpv_list, 
     p_tost_list,
     type="l",
     lwd = 3, 
     ylim = c(0, 1), 
     xlim = c(0, 1),
     # yaxt = "n",
     # xaxt = "n",
     ylab = "TOST p-value",
     xlab = "SGPV")
```

@@

  res <- TOSTone.raw(m = m, 
                     mu = mu,
                     sd = sd, 
                     n = n, 
                     low_eqbound = low_eqbound, 
                     high_eqbound = high_eqbound, 
                     alpha = alpha,
                     plot = FALSE,
                     verbose = FALSE
  )
  t <- (m - mu)/(sd/sqrt(n))
  t_list[count] <- t
  sgpv_list[count] <- p_delta(mu+res$LL_CI_TTEST, mu+res$UL_CI_TTEST, mu+low_eqbound, mu+high_eqbound)
  p_tost_list[count] <- max(res$TOST_p1, res$TOST_p2)
  p_list[count] <- 2 * pt(-abs(t), df = n-1)
}
```
```{r, TOSTSGPV6, fig.cap="Comparison of *p*-values from TOST (black line) and SGPV (dotted grey line) across a range of observed sample means (x-axis). Because the sample size is small (n = 10) and the CI is more than twice as wide as the equivalence range (set to -0.4 to 0.4), the SGPV is set to 0.5 (horizontal lightgrey line) across a range of observed means."}
plot(NA, 
     ylim = c(0, 1), 
     xlim = c(0, 600),
     yaxt = "n",
     xaxt = "n",
     ylab = "P-value for TOST and SGPV",
     xlab = "Observed Mean")
axis(1, at = seq(0,600,100), labels = seq(-3,3,1), las = 1)
axis(2, at = seq(0,1,0.1), labels = seq(0,1,0.1), las = 1)

@@

We can examine this situation by calculating the SGPV and performing the TOST for a situation where sample sizes are small and the equivalence range is narrow, such that the CI is more than twice as large as the equivalence range (see Figure \ref{fig:TOSTSGPV6}). When the two statistics are plotted against each other we can seewhere they are unrelated (indicated by straight lines in the curve, see Figure \ref{fig:TOSTSGPV7}). We see the SGPV is 0.5 for a range of observed means where the *p*-value from the equivalence test still varies. It should be noted that in these calculations the *p*-values for the TOST procedure are *never* smaller than 0.05 (i.e., they do not get below 0.05 on the y-axis). In other words, we cannot conclude equivalence based on any of the observed means. This happens because the TOST *p*-value is smaller than 0.05 only when the 90% CI falls completely between the upper and lower equivalence bounds. However, we are examining a scenario where the 90% CI is so wide that it never falls completely within the two equivalence bounds, and thus the equivalence test is never significant. As @lakens_equivalence_2017 notes: “in small samples (where CIs are wide), a study might have no statistical power (i.e., the CI will always be so wide that it is necessarily wider than the equivalence bounds).” None of the *p*-values based on the TOST procedure are below 0.05, and thus, in the long run we have 0% power.

@@

The *p*-value from the TOST procedure and the SGPV are also unrelated when the CI is wider than the equivalence range (so the precision is low) and overlaps with the upper and lower equivalence bound, but the CI is *not* twice as wide as the equivalence range. In the example below, we see that the CI is only 1.79 times as wide as the equivalence bounds, but the CI overlaps with the lower and upper equivalence bounds (Figure \ref{fig:TOSTSGPV8}). This means the SGPV is not set to 0.5, but it is constant across a range of observed means.
```{r, include=FALSE}
m <- 0.2
n <- 10
sd <- 1
d.lo <- -0.4
d.hi <- 0.4

@@

```
```{r, TOSTSGPV8, fig.cap="Example of a 95% CI that overlaps with the lower and upper equivalence bound (indicated by the vertical dotted lines)."}

@@

If the observed mean would be somewhat closer to 0, or further away from 0, the SGPV remains constant (the CI width does not change, and it completely overlaps with the equivalence range) while the *p*-value for the TOST procedure does vary. We can see this in Figure \ref{fig:TOSTSGPV9} below. The SGPV is not set to 0.5, but is slightly higher than 0.5 across a range of means. How high the SGPV will be for a CI that is not twice as wide as the equivalence range, but overlaps with the lower and upper equivalence bounds, depends on the width of the CI and the equivalence range. 
```{r, include=FALSE}
step = 0.01

@@

  res <- TOSTone.raw(m = m, 
                     mu = mu,
                     sd = sd, 
                     n = n, 
                     low_eqbound = low_eqbound, 
                     high_eqbound = high_eqbound, 
                     alpha = alpha,
                     plot = FALSE,
                     verbose = FALSE
  )
  t <- (m - mu)/(sd/sqrt(n))
  t_list[count] <- t
  sgpv_list[count] <- p_delta(mu+res$LL_CI_TTEST, mu+res$UL_CI_TTEST, mu+low_eqbound, mu+high_eqbound)
  p_tost_list[count] <- max(res$TOST_p1, res$TOST_p2)
  p_list[count] <- 2 * pt(-abs(t), df = n-1)
}
```
```{r, TOSTSGPV9, fig.cap="Comparison of *p*-values from TOST (black line) and SGPV (dotted grey line) across a range of observed sample means (x-axis). The sample size is small (n = 10), but because the sd is half as big as in Figure \\ref{fig:TOSTSGPV7} (1 instead of 2) the CI is less than twice as wide as the equivalence range (set to -0.4 to 0.4). The SGPV is not set to 0.5 (horizontal lightgrey line) but reaches a maximum slightly above 0.5 across a range of observed means."}
plot(NA, 
     ylim = c(0, 1), 
     xlim = c(0, 600),
     yaxt = "n",
     xaxt = "n",
     ylab = "P-value for TOST and SGPV",
     xlab = "Observed Mean")
axis(1, at = seq(0,600,100), labels = seq(-3,3,1), las = 1)
axis(2, at = seq(0,1,0.1), labels = seq(0,1,0.1), las = 1)

@@

@@

If we once more plot the two statistics against each other to see where they are unrelated (indicated by straight lines in the curve), we see the SGPV is 0.56 for a range of observed means where the *p*-value from the equivalence test still varies (Figure \ref{fig:TOSTSGPV10}). To conclude this section, there are three situations where the *p*-value from the TOST procedure is unrelated to the SGPV. In all these situations the *p*-value for the equivalence test differentiates tests with different means, but the SGPV does not. Therefore, as a purely continuous descriptive statistic, the SGPV is more limited than the *p*-value from the TOST procedure.

@@

The effect of asymmetric confidence intervals is most noticable at smaller sample sizes. In Figure \ref{fig:TOSTSGPV11} we plot the *p*-values from equivalence tests and the SGPV (again plotted as 1-SGPV for ease of comparison) for correlations. The sample size is 30 pairs of observations, and the lower and upper equivalence bounds are set to -0.45 and 0.45, with an alpha of 0.05. As the observed correlation in the sample moves from -1 to 0 the *p*-value from the equivalence test becomes smaller, as does 1-SGPV. The pattern is quite similar to that in Figure \ref{fig:TOSTSGPV2}. The *p*-value for the TOST procedure and 1-SGPV are still identical when *p*-values are 0.975 and 0.025 (indicated by the upper and lower horizontal dotted lines). There are two important differences, however. First of all, the SGPV is no longer a straight line, but a curve, due to the asymmetry in the 95% CI. Second, and most importantly, the *p*-value for the equivalence test and the SGPV do no longer overlap at *p* = 0.5.

@@

lines(1-sgpv_list, type="l", col = "darkgrey", lwd = 3, lty = 3)
lines(p_tost_list, lwd = 3)
# abline(h = c(0.975, 0.5, 0.025),
#        lty = 2,
#        col = "grey")

@@

lines(1-sgpv_list, type="l", col = "darkgrey", lwd = 3, lty = 3)
lines(p_tost_list, lwd = 3)
#abline(h = c(0.975, 0.5, 0.025),
#       lty = 2,
#       col = "grey")
```

As a more extreme example of the peculiar behavior of the 'small sample correction' as currently implemented in the calculation of the SGPV see Figure \ref{fig:TOSTSGPV13} below. In this figure observed correlations (from a sample size of 10) from -1 to 1 are tested against an equivalence range from r = 0.4 to r = 0.8. We can see the SGPV has a peculiar shape because it is set to 0.5 for certain observed correlations, even though there is no risk of a misleading SGPV in this range. This example suggests that the current implementation of the 'small sample correction' could be improved. If, on the other hand, the SGPV is mainly meant to be interpreted when it is 0 or 1, it might be preferable to simply never apply the 'small sample correction'.

>>>>

abstract: |
  To move beyond the limitations of null-hypothesis tests, statistical approaches have been developed where the observed data is compared against a range of values that are equivalent to the absence of a meaningful effect. Specifying a range of values around zero allows researchers to statistically reject the presence of effects large enough to matter, and prevents practically insignificant effects from being interpreted as a statistically significant difference. We compare the behavior of the recently proposed second generation *p*-value [@blume_second-generation_2018] with the more established Two One-Sided Tests (TOST) equivalence testing procedure [@schuirmann_comparison_1987]. We show that the two approaches yield almost identical results under optimal conditions. Under suboptimal conditions (e.g., when the confidence interval is wider than the equivalence range, or when confidence intervals are asymmetric) the second generation *p*-value becomes difficult to interpret as a descriptive statistic. The second generation *p*-value is interpretable in a dichotomous manner (i.e., when the SGPV equals 0 or 1 because the confidence intervals lies completely within or outside of the equivalence range), but this dichotomous interpretation does not require calculations. We conclude that equivalence tests yield more consistent *p*-values, distinguish between datasets that yield the same second generation *p*-value, and allow for easier control of Type I and Type II error rates.

@@

The TOST procedure also relies on the confidence interval around the effect. 
In the TOST procedure the data is tested against the lower equivalence bound in the first one-sided test, and against the upper equivalence bound in the second one-sided test [@lakens_equivalence_2018]. 
If both tests statistically reject an effect as extreme or more extreme than the equivalence bound, you can conclude the observed effect is practically equivalent to zero from a Neyman-Pearson approach to statistical inferences. 
Because one-sided tests are performed, one can also conclude equivalence by checking whether the 1-2$\times$$\alpha$ confidence interval (e.g., when the alpha level is 0.05, a 90% CI) falls completely within the equivalence bounds. 
Because both equivalence tests as the SGPV are based on whether and how much a confidence interval overlaps with equivalence bounds, it seems worthwhile to compare the behavior of the newly proposed SGPV to equivalence tests to examine the unique contribution of the SGPV to the statistical toolbox.

@@

The second generation *p*-value (SGPV) is calculated as:
$$
  p _ { \delta } = \frac { \left| I \cap H _ { 0 } \right| } { | I | } \times \max \left\{ \frac { | I | } { 2 \left| H _ { 0 } \right| } , 1 \right\}
$$
where I is the interval based on the data (e.g., a 95% confidence interval) and H~0~ is the equivalence range. The first term of this formula implies that the second generation *p*-value is the width of the confidence interval that overlaps with the equivalence range, divided by the total width of the confidence interval. The second term is a 'small sample correction' (which will be discussed later) that comes into play whenever the confidence interval is more than twice as wide as the equivalence range.
To examine the relation between the TOST *p*-value and the SGPV we can calculate both statistics across a range of observed effect sizes. In Figure \ref{fig:TOSTSGPV1} *p*-values are plotted for the TOST procedure and the SGPV. The statistics are calculated for hypothetical one-sample *t*-tests for observed means ranging from 140 to 150 (on the x-axis). The equivalence range is set to 145 ± 2 (i.e., an equivalence range from 143 to 147), the observed standard deviation is assumed to be 2, and the sample size is 30. For example, for the left-most point in Figure \ref{fig:TOSTSGPV1} the SGPV and the TOST *p*-value is calculated for a hypothetical study with a sample size of 30, an observed standard deviation of 2, and an observed mean of 140, where the *p*-value for the equivalence test is 1, and the SGPV is 0. Our conclusions about the relationship between TOST *p*-values and SGPV in this article are not dependent upon any specific example, as readers can explore for themselves in an online Shiny app: http://shiny.ieis.tue.nl/TOST_vs_SGPV/.

@@

tost_res2 <- TOSTone.raw(m = m, 
                   mu = mu,
                   sd = sd, 
                   n = n, 
                   low_eqbound = low_eqbound, 
                   high_eqbound = high_eqbound, 
                   alpha = alpha,
                   plot = FALSE,
                   verbose = FALSE
)
SGPV2 <- TOST_to_SGPV(tost_res = tost_res2)
```

When the observed sample mean is 145, the sample size is 30, and the standard deviation is 2, and we are testing against equivalence bounds of 143 and 147 using the TOST procedure for a one-sample *t*-test, the equivalence test is significant, *t*(`r tost_res1$TOST_df`) = `r round(tost_res1$TOST_t1, digits = 2)`, *p* < .001. Because the 95% CI falls completely within the equivalence bounds, the SGPV is `r SGPV1` (see Figure \ref{fig:TOSTSGPV1}). 
On the other hand, when the observed mean is 140, the equivalence test is not significant (the observed mean is far outside the equivalence range of 143 to 147), *t*(`r tost_res2$TOST_df`) = `r round(tost_res2$TOST_t1, digits = 2)`, *p* = `r round(tost_res2$TOST_p1, digits =  4)` (or more accurately, *p* > .999 as *p*-values are bounded between 0 and 1). Because the 95% CI falls completely outside the equivalence bounds, the SGPV is `r SGPV2` (see Figure  \ref{fig:TOSTSGPV1}).

@@

Two other points always have to overlap. When the 95% CI falls completely inside the equivalence region, and one endpoint of the confidence interval is exactly equal to one of the equivalence bounds (see situation B in Figure \ref{fig:TOSTSGPV3}) the TOST *p*-value (which relies on a one-sided test) is always 0.025. The third point where the SGPV and the *p*-value from the TOST procedure should overlap is where the 95% CI falls completely outside of the equivalence range, but one endpoint of the confidence interval is equal to the equivalence bound (see situation C in Figure \ref{fig:TOSTSGPV3}), when the *p*-value will always be 0.975. Note that this situation is in essence a minimum-effect test [@murphy_statistical_2014]. The goal of a minimum-effect is not just to reject a difference of zero, but to reject the smallest effect size of interest (i.e., the equivalence bounds). The SGPV summarizes the information from an equivalence test and a minimum-effect test. These can be two relevant questions to ask, although it often makes sense to combine an equivalence test and a null-hypothesis test instead [@lakens_equivalence_2018].

@@

tost_res4 <- TOSTone.raw(m = m, 
                   mu = mu,
                   sd = sd, 
                   n = n, 
                   low_eqbound = low_eqbound, 
                   high_eqbound = high_eqbound, 
                   alpha = alpha,
                   plot = FALSE,
                   verbose = FALSE
)
SGPV4 <- TOST_to_SGPV(tost_res = tost_res4)
```
```{r , echo=FALSE}
tail1 <- 1-pnorm(2, 1.5, 0.5)
tail2 <- 1-pnorm(2, 1.4, 0.5)
tail3 <- 1-pnorm(2, 1.3, 0.5)
tail4 <- 1-pnorm(2, 1.2, 0.5)
```

For example, the SGPV from A to D is `r round(SGPV1, digits = 2)`, `r round(SGPV2, digits = 2)`, `r round(SGPV3, digits = 2)`, and `r round(SGPV4, digits = 2)`. The difference in the percentage of overlap between A and B (`r round(SGPV1-SGPV2, digits = 2)`) is identical to the difference in the percentage of overlap between C and D as the mean gets 0.1 closer to the test value (`r round(SGPV3-SGPV4, digits = 2)`).
As the observed mean in a one-sample *t*-test lies closer to the test value (in Figure \ref{fig:TOSTSGPV4}), from situation A to D, the mean gets closer to the test value by 0.1) the difference in the overlap changes uniformly. 
As we move the observed mean closer to the test value in steps of 0.1 across A to D the *p*-value calculated for normally distributed data is not uniformly distributed. The probability of observing data more extreme than the upper bound of 2 is (from A to D) `r round(tail1, digits = 3)`, `r round(tail2, digits = 3)`, `r round(tail3, digits = 3)`, and `r round(tail4, digits = 3)`. As we can see, the difference between A and B (`r round(tail1-tail2, digits = 3)`) is not the same as the difference between C and D (`r round(tail3-tail4, digits = 3)`). Indeed, the difference in *p*-values is the largest as you start at *p* = 0.5 (when the observed mean falls on the test value), which is why the line in Figure \ref{fig:TOSTSGPV1} is the steepest at *p* = 0.5. Note that where the SGPV reaches 1 or 0, *p*-values closely approximate 0 and 1, but never reach these values.

@@

There are three situations where *p*-values from TOST and SGPV are unrelated. The first two situations were discussed before and can be seen in Figure \ref{fig:TOSTSGPV1}. When the SGPV is either 0 or 1 *p*-values from the equivalence test fall between 0.975 and 1 or between 0 and 0.025. Because *p*-values approach 0 or 1, but are never exactly 0 or 1, and the SGPV is exactly 0 or 1, the two statistics are completely unrelated within these ranges. The easiest way to see this is by plotting the SGPV against the *p*-value from the TOST procedure. The situations where the SPGV and *p*-values from the TOST procedure are unrelated are indicated by the parts of the curve where there are vertical straight lines at second generation *p*-values of 0 and 1.

@@

  res <- TOSTone.raw(m = m, 
                     mu = mu,
                     sd = sd, 
                     n = n, 
                     low_eqbound = low_eqbound, 
                     high_eqbound = high_eqbound, 
                     alpha = alpha,
                     plot = FALSE,
                     verbose = FALSE
  )
  t <- (m - mu)/(sd/sqrt(n))
  t_list[count] <- t
  sgpv_list[count] <- p_delta(mu+res$LL_CI_TTEST, mu+res$UL_CI_TTEST, mu+low_eqbound, mu+high_eqbound)
  p_tost_list[count] <- max(res$TOST_p1, res$TOST_p2)
  p_list[count] <- 2 * pt(-abs(t), df = n-1)
}
```

```{r, TOSTSGPV5, fig.cap="The relationship between *p*-values from the TOST procedure and the SGPV for the same scenario as in Figure \\ref{fig:TOSTSGPV1}."}
plot(sgpv_list, 
     p_tost_list,
     type="l",
     lwd = 3, 
     ylim = c(0, 1), 
     xlim = c(0, 1),
     # yaxt = "n",
     # xaxt = "n",
     ylab = "TOST p-value",
     xlab = "SGPV")
```

@@

  res <- TOSTone.raw(m = m, 
                     mu = mu,
                     sd = sd, 
                     n = n, 
                     low_eqbound = low_eqbound, 
                     high_eqbound = high_eqbound, 
                     alpha = alpha,
                     plot = FALSE,
                     verbose = FALSE
  )
  t <- (m - mu)/(sd/sqrt(n))
  t_list[count] <- t
  sgpv_list[count] <- p_delta(mu+res$LL_CI_TTEST, mu+res$UL_CI_TTEST, mu+low_eqbound, mu+high_eqbound)
  p_tost_list[count] <- max(res$TOST_p1, res$TOST_p2)
  p_list[count] <- 2 * pt(-abs(t), df = n-1)
}
```

```{r, TOSTSGPV6, fig.cap="Comparison of *p*-values from TOST (black line) and SGPV (dotted grey line) across a range of observed sample means (x-axis). Because the sample size is small (n = 10) and the CI is more than twice as wide as the equivalence range (set to -0.4 to 0.4), the SGPV is set to 0.5 (horizontal light grey line) across a range of observed means."}
plot(NA, 
     ylim = c(0, 1), 
     xlim = c(0, 600),
     yaxt = "n",
     xaxt = "n",
     ylab = "P-value for TOST and SGPV",
     xlab = "Observed Mean")
axis(1, at = seq(0,600,100), labels = seq(-3,3,1), las = 1)
axis(2, at = seq(0,1,0.1), labels = seq(0,1,0.1), las = 1)

@@

We can examine this situation by calculating the SGPV and performing the TOST for a situation where sample sizes are small and the equivalence range is narrow, such that the CI is more than twice as large as the equivalence range (see Figure \ref{fig:TOSTSGPV6}). When the two statistics are plotted against each other we can see where they are unrelated (indicated by straight lines in the curve, see Figure \ref{fig:TOSTSGPV7}). We see the SGPV is 0.5 for a range of observed means where the *p*-value from the equivalence test still varies. It should be noted that in these calculations the *p*-values for the TOST procedure are *never* smaller than 0.05 (i.e., they do not get below 0.05 on the y-axis). In other words, we cannot conclude equivalence based on any of the observed means. This happens because the TOST *p*-value is smaller than 0.05 only when the 90% CI falls completely between the upper and lower equivalence bounds. However, we are examining a scenario where the 90% CI is so wide that it never falls completely within the two equivalence bounds, and thus the equivalence test is never significant. As @lakens_equivalence_2017 notes: “in small samples (where CIs are wide), a study might have no statistical power (i.e., the CI will always be so wide that it is necessarily wider than the equivalence bounds).” None of the *p*-values based on the TOST procedure are below 0.05, and thus, in the long run we have 0% power.

@@

The *p*-value from the TOST procedure and the SGPV are also unrelated when the CI is wider than the equivalence range (so the precision is low) and overlaps with the upper and lower equivalence bound, but the CI is *not* twice as wide as the equivalence range. In the example below, we see that the CI is only 1.79 times as wide as the equivalence bounds, but the CI overlaps with the lower and upper equivalence bounds (Figure \ref{fig:TOSTSGPV8}). This means the SGPV is not set to 0.5, but it is constant across a range of observed means.

```{r, include=FALSE}
m <- 0.2
n <- 10
sd <- 1
d.lo <- -0.4
d.hi <- 0.4

@@

```{r, TOSTSGPV8, fig.cap="Example of a 95% CI that overlaps with the lower and upper equivalence bound (indicated by the vertical dotted lines)."}

@@

If the observed mean would be somewhat closer to 0, or further away from 0, the SGPV remains constant (the CI width does not change, and it completely overlaps with the equivalence range) while the *p*-value for the TOST procedure does vary. We can see this in Figure \ref{fig:TOSTSGPV9} below. The SGPV is not set to 0.5, but is slightly higher than 0.5 across a range of means. How high the SGPV will be for a CI that is not twice as wide as the equivalence range, but overlaps with the lower and upper equivalence bounds, depends on the width of the CI and the equivalence range.

@@

  res <- TOSTone.raw(m = m, 
                     mu = mu,
                     sd = sd, 
                     n = n, 
                     low_eqbound = low_eqbound, 
                     high_eqbound = high_eqbound, 
                     alpha = alpha,
                     plot = FALSE,
                     verbose = FALSE
  )
  t <- (m - mu)/(sd/sqrt(n))
  t_list[count] <- t
  sgpv_list[count] <- p_delta(mu+res$LL_CI_TTEST, mu+res$UL_CI_TTEST, mu+low_eqbound, mu+high_eqbound)
  p_tost_list[count] <- max(res$TOST_p1, res$TOST_p2)
  p_list[count] <- 2 * pt(-abs(t), df = n-1)
}
```

```{r, TOSTSGPV9, fig.cap="Comparison of *p*-values from TOST (black line) and SGPV (dotted grey line) across a range of observed sample means (x-axis). The sample size is small (n = 10), but because the sd is half as big as in Figure \\ref{fig:TOSTSGPV7} (1 instead of 2) the CI is less than twice as wide as the equivalence range (set to -0.4 to 0.4). The SGPV is not set to 0.5 (horizontal light grey line) but reaches a maximum slightly above 0.5 across a range of observed means."}
plot(NA, 
     ylim = c(0, 1), 
     xlim = c(0, 600),
     yaxt = "n",
     xaxt = "n",
     ylab = "P-value for TOST and SGPV",
     xlab = "Observed Mean")
axis(1, at = seq(0,600,100), labels = seq(-3,3,1), las = 1)
axis(2, at = seq(0,1,0.1), labels = seq(0,1,0.1), las = 1)

@@

If we once more plot the two statistics against each other to see where they are unrelated (indicated by straight lines in the curve), we see the SGPV is 0.56 for a range of observed means where the *p*-value from the equivalence test still varies (Figure \ref{fig:TOSTSGPV10}).

@@

To conclude this section, there are three situations where the *p*-value from the TOST procedure is unrelated to the SGPV. In all these situations the *p*-value for the equivalence test differentiates tests with different means, but the SGPV does not. Therefore, as a purely continuous descriptive statistic, the SGPV is more limited than the *p*-value from the TOST procedure.

@@

The effect of asymmetric confidence intervals is most noticeable at smaller sample sizes. In Figure \ref{fig:TOSTSGPV11} we plot the *p*-values from equivalence tests and the SGPV (again plotted as 1-SGPV for ease of comparison) for correlations. The sample size is 30 pairs of observations, and the lower and upper equivalence bounds are set to -0.45 and 0.45, with an alpha of 0.05. As the observed correlation in the sample moves from -1 to 0 the *p*-value from the equivalence test becomes smaller, as does 1-SGPV. The pattern is quite similar to that in Figure \ref{fig:TOSTSGPV2}. The *p*-value for the TOST procedure and 1-SGPV are still identical when *p*-values are 0.975 and 0.025 (indicated by the upper and lower horizontal dotted lines). There are two important differences, however. First of all, the SGPV is no longer a straight line, but a curve, due to the asymmetry in the 95% CI. Second, and most importantly, the *p*-value for the equivalence test and the SGPV do no longer overlap at *p* = 0.5.

@@

lines(1-sgpv_list, type="l", col = "darkgrey", lwd = 3, lty = 3)
lines(p_tost_list, lwd = 3)
abline(h = c(0.975, 0.5, 0.025),
       lty = 2,
       col = "grey")

@@

lines(1-sgpv_list, type="l", col = "darkgrey", lwd = 3, lty = 3)
lines(p_tost_list, lwd = 3)
abline(h = c(0.975, 0.5, 0.025),
       lty = 2,
       col = "grey")
```

As a more extreme example of the peculiar behavior of the 'small sample correction' as currently implemented in the calculation of the SGPV see Figure \ref{fig:TOSTSGPV13} below. In this figure observed correlations (from a sample size of 10) from -1 to 1 are tested against an equivalence range from r = 0.4 to r = 0.8. We can see the SGPV has a peculiar shape because it is set to 0.5 for certain observed correlations, even though there is no risk of a 'misleading' SGPV in this range. This example suggests that the current implementation of the 'small sample correction' could be improved. If, on the other hand, the SGPV is mainly meant to be interpreted when it is 0 or 1, it might be preferable to simply never apply the 'small sample correction'.
